Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-30 _tex/paper2.tex
Built site for gh-pages

\caption{\label{fig-caseV_issues}The effect of dispersion discrepancies
and stimulus correlation on the distribution of the discriminal
difference.}

@@

Before integrating the theoretical principles from
Section~\ref{sec-thurstone_theory}, it is important to define SCMs and
DAGs. SCMs are formal mathematical models defined by a set of
\emph{endogenous} variables, a set of \emph{exogenous} variables, and a
set of functions \citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous
variables are those whose causal mechanisms a researcher chooses to
model \citep{Neal_2020}. In contrast, exogenous variables represent
\emph{errors} or \emph{disturbances} arising from omitted factors that
the investigator chooses not to model explicitly
\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
\emph{structural equations}, express the endogenous variables as
non-parametric functions of other variables.

For instance, Figure~\ref{fig-CJ1_scm} display the SCM relating the CJ
outcome \((O_{kjabi_{1}i_{2}})\), discriminal difference
\((D_{kjabi_{1}i_{2}})\), perceived discriminal processes for each text
within each student \((PT_{kjai_{1}}, PT_{kjbi_{2}})\), the ``true''
discriminal processes for each text within each student
\((T_{kjai_{1}}, T_{kjbi_{2}})\), and the judges' biases \((B_{kj})\).
Notice that the structural equations use the symbol `\(:=\)' to denote
the asymmetrical causal dependence between variables and the symbol
`\(\:\bot\:\)' to represent \emph{d-separation}, a concept akin to
(conditional) independence.

@@

\[
\begin{aligned}
  O_{kjabi_{1}i_{2}} & := f_{O}(D_{kjabi_{1}i_{2}}) \\
  D_{kjabi_{1}i_{2}} & := f_{D}(PT_{kjai_{1}}, PT_{kjbi_{2}}) \\
  PT_{kjai_{1}} & := f_{PT}(T_{ai_{1}}, T_{bi_{2}}, B_{kj})
\end{aligned}
\]

@@

\subcaption{\label{fig-CJ1_scm}SCM}

@@

\subcaption{\label{fig-CJ1_DAG}DAG}

@@

\caption{\label{fig-CJ1}CJ's outcome, discriminal difference,
discriminal processes and judges' biases}

@@

without loosing generality, the (latent) ``perceived'' and ``true''
discriminal processes for the stimuli can be depicted in a vector for
each judge, as in

@@

\caption{\label{fig-CJ_TM_05}}

@@

@@

\caption{\label{fig-CJ_TM_10}}

@@

@@

\caption{\label{fig-CJ_TM_12}}

@@

@@

\caption{\label{fig-CJ_TM_14}}

@@

@@

\caption{\label{fig-CJ_TM_15}}

>>>>

\caption{\label{fig-casev_issues}The effect of dispersion discrepancies
and stimulus correlation on the distribution of the discriminal
difference.}

@@

Before integrating the theoretical principles from
Section~\ref{sec-thurstone_theory}, it is important to define SCMs and
DAGs. SCMs are formal mathematical models defined by a set of
\emph{endogenous} variables, a set of \emph{exogenous} variables, and a
set of functions \citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous
variables are those whose causal mechanisms a researcher chooses to
model \citep{Neal_2020}. In contrast, exogenous variables represent
\emph{errors} or \emph{disturbances} arising from omitted factors that
the investigator chooses not to model explicitly
\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
\emph{structural equations}, express the endogenous variables as
non-parametric functions of other variables. These functions use the
symbol `\(:=\)' to denote the asymmetrical causal dependence between
variables and the symbol `\(\:\bot\:\)' to represent
\emph{d-separation}, a concept akin to (conditional) independence.

For instance, following the scenario outline in
Section~\ref{sec-theoretical}, Figure~\ref{fig-cj04_scm} displays the
SCM that relates the outcome of the CJ experiment \((O_{abhijk})\), to
several endogenous variables \(V\). These variables include the
discriminal difference \((D_{abhijk})\), perceived discriminal processes
\((PT_{ahjk}, PT_{bijk})\), the ``true'' discriminal processes
\((T_{ahjk}, T_{bijk})\), and the judges' biases \((B_{kj})\), formally
expressed as
\(V = \{D_{abhijk}, PT_{ahjk}, PT_{bijk}, T_{ahjk}, T_{bijk}, B_{kj}\}\).
The structural equations \(\{f_{O},f_{D},f_{PT}\}\) define the
dependencies among these variables. Notably, in this scenario, the
judges produce a dichotomous outcome \((O_{abhijk}) = \{0,1\}\).
Moreover, the indices \(\{a,b\}\) represent the compared texts,
\(\{h,i\}\) label the students who wrote these texts, \(j\) denotes the
judge, and \(k\) represents the comparison number. Regarding the \(k\)
index, the index accounts for cases in which judges perform multiple
comparisons of the same texts written by the same students, as the one
discussed in Section~\ref{sec-theoretical}.

@@

\[
\begin{aligned}
  O_{abhijk} & := f_{O}(D_{abhijk}) \\
  D_{abhijk} & := f_{D}(PT_{ahjk}, PT_{bijk}) \\
  PT_{ahjk} & := f_{PT}(T_{ah}, B_{jk}) \\
  PT_{bijk} & := f_{PT}(T_{bi}, B_{jk})
\end{aligned}
\]

@@

\subcaption{\label{fig-cj04_scm}SCM}

@@

\subcaption{\label{fig-cj04_dag}DAG}

@@

\caption{\label{fig-cj04}CJ's outcome, discriminal difference,
discriminal processes and judges' biases}

@@

Because it quickly becomes intractable, and without loosing generality,
the (latent) ``perceived'' and ``true'' discriminal processes for the
stimuli can be depicted in a vector for each judge, as in

\[
\begin{aligned}
  O_{PRQ} & := f_{O}(D_{PRQ}) \\
  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ})
\end{aligned}
\]

\subcaption{\label{fig-cj05_scm}SCM}

@@

\subcaption{\label{fig-cj05_dag}DAG}

\caption{\label{fig-cj05}CJ's outcome, discriminal difference,
discriminal processes and judges' biases}

@@

\[
\begin{aligned}
  O_{PRQ} & := f_{O}(D_{PRQ}) \\
  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) 
\end{aligned}
\]

\subcaption{\label{fig-cj10_scm}SCM}

@@

\subcaption{\label{fig-cj10_dag}DAG}

\caption{\label{fig-cj10}CJ's outcome, discriminal difference,
discriminal processes and judges' biases}

@@

\[
\begin{aligned}
  O^{S}_{PRQ} & := f_{O}(O_{PRQ}, S_{I}, S_{A}, S_{K}, S_{J}) \\
  O_{PRQ} & := f_{O}(D_{PRQ}) \\
  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J})
\end{aligned}
\]

\subcaption{\label{fig-cj12_scm}SCM}

@@

\subcaption{\label{fig-cj12_dag}DAG}

\caption{\label{fig-cj12}CJ's outcome, discriminal difference,
discriminal processes and judges' biases}

@@

\[
\begin{aligned}
  O^{C}_{PRQ} & := f_{C}(O^{S}_{PRQ}, C) \\
  O^{S}_{PRQ} & := f_{S}(O_{PRQ}, S_{I}, S_{A}, S_{K}, S_{J}) \\
  O_{PRQ} & := f_{O}(D_{PRQ}) \\
  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) 
\end{aligned}
\]

\subcaption{\label{fig-cj14_scm}SCM}

@@

\subcaption{\label{fig-cj14_dag}DAG}

\caption{\label{fig-cj14}CJ's outcome, discriminal difference,
discriminal processes and judges' biases}

@@

\[
\begin{aligned}
  O^{C}_{PRQ} & := f_{C}(O^{S}_{PRQ}, C) \\
  O^{S}_{PRQ} & := f_{S}(O_{PRQ}, S_{I}, S_{A}, S_{K}, S_{J}) \\
  O_{PRQ} & := f_{O}(D_{PRQ}) \\
  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) 
\end{aligned}
\]

\subcaption{\label{fig-cj15_scm}SCM}

@@

\subcaption{\label{fig-cj15_dag}DAG}

\caption{\label{fig-cj15}CJ's outcome, discriminal difference,
discriminal processes and judges' biases}
